---
title: "initial_model"
output: html_document
date: "2024-03-14"
---

## Load Libraries
```{r}
library(caret)
library(corrplot)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(glmnet)
library(tidyverse)
library(MASS)
library(pROC)
library(Information)
library(performanceEstimation)
```


#Exploratory Data Analysis

## Read in data
```{r}
# Read data
census_data <- read.csv("../Data/census_data.csv", header=TRUE, na.strings=c(c("", ' ?'),"NA"))

# Drop ID column
df <- subset(census_data, select = -c(1))
```

## Percent missing by column
```{r, echo=FALSE}
# Percent missing ----------------------------------------------
pct_missing <- function(data) {
  # Calculate the total number of observations
  total_obs <- nrow(data)
  
  # Calculate the number of missing values in each column
  missing_count <- colSums(is.na(data))
  
  # Calculate the percentage of missing values in each column
  percent_missing <- (missing_count / total_obs) *100
  
  # Create a data frame to store the results
  missing_data <- data.frame(Column = names(data), Percent_Missing = percent_missing)
  
  # Return the data frame
  return(missing_data)
}

percent_missing <- pct_missing(df)
print(percent_missing)
```

## Columns to drop
```{r, echo=FALSE}
# Drop columns with missing values over 90% of the time
over_90 <- percent_missing$Percent_Missing > 90
drop_missing <- percent_missing[over_90, ]

# Columns to drop
drop_missing$Column

# Variance ---------------------------------------------------- 
near_zero_variance <- function(data) {

    near_zero_vars <- nearZeroVar(data, names = TRUE, saveMetrics = TRUE)
    
    # Create a data frame to store the results
    variance_data <- data.frame(Column = names(data), ZeroVariance = near_zero_vars$zeroVar, NearZeroVariance = near_zero_vars$nzv)
    
    return(variance_data)
}

near_zero_vars <- near_zero_variance(df)

# Drop columns with zero variance
zero_var <- near_zero_vars$ZeroVariance == TRUE
drop_variance <- near_zero_vars[zero_var, ]

# Columns to drop
drop_variance$Column

# Drop columns from above analysis
df <- subset(df, select = -c(education_institute, unemployment_reason, 
                                       under_18_family, migration_prev_sunbelt,
                                       veterans_admin_questionnaire, 
                                       old_residence_reg, old_residence_state))
```

#Variable Selection

## Clean up census_data based on EDA
```{r}
set.seed(1234)

#Drop columns discovered in EDA.R
drop <- match(c("ID", "education_institute", "under_18_family", "veterans_admin_questionnaire", 
        "old_residence_reg", "old_residence_state", "unemployment_reason", "migration_prev_sunbelt"), names(census_data))
df <- subset(census_data, select = -drop)

#Remove whitespace from column data
df <- df %>%
  mutate(across(where(is.character), str_trim))

#convert non-numeric columns to factors
cols = c("age", "employment_stat", "wage_per_hour", "working_week_per_year", "industry_code", 
         "occupation_code", "total_employed", "gains", "losses", "stocks_status", "mig_year", "importance_of_record")
df[,!names(df) %in% cols] <- lapply(df[,!names(df) %in% cols], as.factor)

#Drop Columns with Na proportion >20%
df <- df %>%
  select_if(~mean(is.na(.)) < .2)

#Impute missing values w/ simple Mode imputation for factor items
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
df <- df %>%
  mutate_if(is.factor, list(~replace(., is.na(.), mode(na.omit(.)))))

#Proportions of NA data after cleanup
df %>%
  mutate(across(everything(), is.na)) %>%
  arrange(across(everything())) %>%
  mutate(row = row_number()) %>%
  pivot_longer(!row, names_to="column", values_to="missing") %>%
  ggplot() +
  geom_tile(aes(row, column, fill=missing)) + ggtitle("Missing Observation Occurances") + xlab("Observation") + ylab("Variable")
```
## Explore outliers + Data Cleaning
```{r}
numeric_cols <- df[sapply(df, is.numeric)]
str(numeric_cols)
#cor(numeric_cols) # Initial findings is there is little linear correlation between numeric variables.
``` 
```{r}
# Replace 'education' (16 levels) with 'education_grp': low_ed, mid_ed, high_ed
df <- df %>%
  mutate(education_grp = case_when(
    education %in% c("Less than 1st grade", "1st 2nd 3rd or 4th grade", "5th or 6th grade", "7th and 8th grade")
    ~ "low_ed",
    education %in% c("9th grade", "10th grade", "11th grade", "High school graduate",  " 12th grade no diploma" ) 
    ~ "mid_ed",
    TRUE ~ "high_ed"))
df$education_grp <- factor(df$education_grp, levels = c("high_ed", "mid_ed", "low_ed"))

# Replace 'class' (4 levels) with 'govt_employee': yes or no
df <- df %>%
  mutate(govt_employee = 
           ifelse(class %in% c("Federal government", "Local government", "State government"), 
                  "yes", "no"))
df$govt_employee <- factor(df$govt_employee, levels = c("yes", "no"))

# Replace 'marital_status' (7 levels) with 'is_married': yes or no
# Include "married- spouse absent' with 'is_married: no'. 
df <- df %>%
  mutate(is_married = 
           ifelse(marital_status %in% c("Married-civilian spouse present", "Married-A F spouse present,
                                        "), "yes", "no"))
df$is_married <- factor(df$is_married, levels = c("yes", "no"))

# Replace 'is_hispanic' (10 levels) with 'is_hispanic_grp': yes, no, or unknown
df <- df %>%
  mutate(is_hispanic_grp = case_when(
    is_hispanic %in% c("Mexican-American", "Mexican (Mexicano)", "Chicano", 
                       "Central or South American", "Puerto Rican", "Other Spanish", "Cuban") ~ "yes",
    is_hispanic %in% c("All other") ~ "no",
    is_hispanic %in% c("NA", "Do not know") ~ "unknown"))
df$is_hispanic_grp <- factor(df$is_hispanic_grp, levels = c("yes", "no", "unknown"))

# Replace 'employment_commitment' (7 levels) with 'work_hours': full-time, part-time, or unemployed
df <- df %>%
  mutate(work_hours =
           ifelse(employment_commitment %in% c("Full-time schedules", 
                                               "Children or Armed Forces"),
           "full-time",
           ifelse(employment_commitment %in% c("PT for non-econ reasons usually FT", 
                                               "PT for econ reasons usually FT", 
                                               "Unemployed full-time"),
           "part-time", "unemployed")))
df$work_hours <- factor(df$work_hours, levels = c("full-time", "part-time", "unemployed"))

# Replace 'tax_status' (6 levels) with 'tax_status_grp': single, joint, head-of-household, nonfiler, or unknown
df <- df %>%
  mutate(tax_status_grp = 
           ifelse(tax_status %in% c("Joint both under 65", 
                                    "Joint one under 65 & one 65+", 
                                    "Joint both 65+"),
                  "joint",
                  ifelse(tax_status == "Single",
                         "single",
                         ifelse(tax_status == "Head of household",
                                "head-of-household",
                                ifelse(tax_status == "Nonfiler",
                                       "nonfiler", "unknown")))))
df$tax_status_grp <- factor(df$tax_status_grp, levels = 
                              c("single", "joint", "head-of-household", "nonfiler", "unknown"))
         
# Replace 'citizenship' (5 levels) with 'citizenship_grp': citizen, non-citizen, or naturalized citizen
df <- df %>%
  mutate(citizenship_grp = case_when(
    citizenship %in% c("Native", "Native- Born in Puerto Rico or U S Outlying", 
                       "Native- Born abroad of American Parent(s)") ~ "citizen",
    citizenship %in% "Foreign born- Not a citizen of U S " ~ "non-citizen",
    citizenship %in% "Foreign born- U S citizen by naturalization" ~ "naturalized citizen")) 
df$citizenship_grp <- factor(df$citizenship_grp, levels = 
                              c("citizen", "non-citizen", "naturalized citizen"))

# For 'is_labor_union, replace "Same" with "Yes"
#df <- df %>%
#  mutate(is_labor_union = factor(
#    ifelse(is_labor_union == "Same", "Yes", is_labor_union),
#    levels = c("yes", "no") ))
```

```{r}
hist(df$age, xlab = "Age", main = "Age Frequency") # skewed slightly right, might want to group into bands for higher age groups.
summary(df$age) 
table(df$age)

df$age_grp <- as.factor(ifelse(df$age <25, "<25", ifelse(df$age <30, "25-29", 
      ifelse(df$age <40, "30-39", ifelse(df$age <50, "40-49", ifelse(df$age <60, "50-59", "60+"))))))
table(df$age_grp) # lowest group frequency is 1126 for 60+. Use age_grp variable

hist(df$employment_stat, xlab = "Employment Stat", main = "Employment Stat Frequency")
table(df$employment_stat) # this is a categorical factor
df$employment_stat <- as.factor(df$employment_stat)

hist(df$wage_per_hour, xlab = "Wage per Hour", main = "Wage per Hour Frequency") # examine for outliers
summary(df$wage_per_hour) # Minimum 0 and Maximum 9999, manipulate?
length(which(df$wage_per_hour > 2000)) # consider dropping or grouping due to large number of outliers (~537)
df$wage_grps <- as.factor(ifelse(df$wage_per_hour == 0, "0", ifelse(df$wage_per_hour < 1000, "1-999", 
      ifelse(df$wage_per_hour < 2000, "1000-1999","2000+"))))
table(df$wage_grps) # lowest group frequency is 643 for 2000+. Use wage_grps variable.

hist(df$working_week_per_year, xlab = "Working Week per Year", main = "Working Week per Year Frequency")
summary(df$working_week_per_year)
table(df$working_week_per_year) # decent amount of observations in most weeks
df$working_weeks_grp <- as.factor(ifelse(df$working_week_per_year < 27, "0-26", ifelse(df$working_week_per_year <52, "27-51", "52")))
table(df$working_weeks_grp) # lowest group frequency is 687 for less than half the year. Use working_weeks_grp variable.

hist(df$industry_code, xlab = "Industry Code", main = "Industry Code Frequency")
summary(df$industry_code)
table(df$industry_code) # lowest number of observations in a group is 9 (code 20), rest are 25+
df$industry_code <- as.factor(df$industry_code) 

hist(df$occupation_code, xlab = "Occupation Code", main = "Occupation Code Frequency") # look at highest value
summary(df$occupation_code)
table(df$occupation_code) # lowest number of observations in a group is 7 (code 20), rest are 19+
df$occupation_code <- as.factor(df$occupation_code)

hist(df$total_employed, xlab = "Total Employed", main = "Total Employed Frequency") # some are 0?
summary(df$total_employed)
table(df$total_employed) # group of 0 has 384 observations

hist(df$gains, xlab = "Gains", main = "Gains Frequency") # examine for outliers
summary(df$gains)
table(df$gains) # recommend dropping. Due to large amount of 0.

hist(df$losses, xlab = "Losses", main = "Losses Frequency") # examine for outliers
summary(df$losses)
table(df$losses) # recommend dropping. Due to large amount of 0. 

hist(df$stocks_status, xlab = "Stock Status", main = "Stock Status Frequency") # examine for outliers
summary(df$stocks_status)
#table(df$stocks_status) # recommend dropping or grouping into bins
df$stock_status_grp <- as.factor(ifelse(df$stocks_status == 0, "0", ifelse(df$stocks_status < 1000, "1-999", "1000+")))
table(df$stock_status_grp) # not sure how useful this is.

hist(df$mig_year, xlab = "Mig Year", main = "Mig Year Frequency") # examine for outliers
summary(df$mig_year)
table(df$mig_year) # categorical 94 or 95, fine to use.
df$mig_year <- as.factor(df$mig_year)

hist(df$importance_of_record, xlab = "Importance of Record", main = "Importance of Record Frequency") # examine for outliers
summary(df$importance_of_record)
#table(df$importance_of_record) # don't use or split into bins.
df$importance_grp <- as.factor(ifelse(df$importance_of_record <1000, "<1000", ifelse(df$importance_of_record <2000, "1000-1999",
      ifelse(df$importance_of_record <3000, "2000-2999", "3000+"))))
table(df$importance_grp) # use this grouped variable.

## Summary of outliers:

# Use age_grp instead of age, use wage_grps instead of wage_per_hour, 
# use importance_grp instead of importance_of_record.

# Turned Employment Stat into factor. Turned industry code into a factor. Turned occupation code into a factor. Can we group codes further?
# Turned Mig Year into factor.

# Recommend dropping gains and losses. Possible stocks_status, or use stock_status_grp.
```

# Drop columns from above analysis
```{r}

# Drop from df: 'age', 'wage_per_hour', 'importance_of_record', 'stocks_status', 'working_week_per_year' 
# 'education', 'class', 'marital_status', 'is_hispanic', 'employment_commitment', 'tax_status', 
# 'citizenship'
df = subset(df, select = -c(age, wage_per_hour, importance_of_record, stocks_status, working_week_per_year,
                            education, class, marital_status, is_hispanic, employment_commitment,
                            tax_status, citizenship))

# Instead use: 'age_grp', 'wage_grp', 'importance_grp', 'stock_status_grp', 'working_weeks_grp'
# 'education_grp, 'govt_employee', 'is_married', 'is_hispanic_grp', 'work_hours', 'tax_status_grp',
# 'citizenship_grp'

# drop 'gains' and 'losses' per above recommendation (large amount of 0's)
df = subset(df, select = -c(gains, losses))

str(df)
```

## Split into training and testing data
```{r}
sample <- sample(c(TRUE, FALSE), nrow(df), replace=TRUE, prob=c(0.7,0.3))
train  <- df[sample, ]
test   <- df[!sample, ]
```
## Create lasso and run cv glmnet
```{r}
#LASSO - separate into feature and target set
y <- train$is_labor_union
X <- data.matrix(train[, !names(train) %in% c("is_labor_union")])
X[is.na(X)] <- 2

#Create cv sets at 10 folds, models, and coefficients and select for 20 variables 
lasso_cv <- cv.glmnet(X, y, alpha=1, nfolds=10, family="binomial", type.measure = "mse", dfmax=20)
lasso_model <- glmnet(X, y, alpha=1, nlambda=100, family = "binomial", dfmax=20)
lasso_coef <- coef(lasso_model, s=lasso_cv$lambda.min, family="binomial")
```

## Lasso model glmnet improvement
```{r, echo=FALSE}
#Plot model and its minimum lambda found
plot(lasso_cv)
```

## Best glmnet model lambda
```{r}
plot(lasso_model, xvar="lambda", lwd=2)
abline(v=log(lasso_cv$lambda.min), col='black', lty=2, lwd=2)


#Model Coefficients
lasso_coef
coefs = as.data.frame(as.matrix(lasso_coef))
zeros <- coefs %>%
  filter(s1 == 0)
coefs <- coefs %>% 
  filter(s1 != 0)
```

## Final feature selection
### Not selected
```{r}
#Features not selected with threshold 20
zeros
```
### Selected
```{r}
#Features selected with threshold 20
coefs
```

## Drop the variables forced to zero by lasso
```{r}
df <- df[,!(names(df) %in% c("household_stat","household_summary","vet_benefit",
                             "country_of_birth_own", "country_of_birth_father", 
                             "is_hispanic_grp", "work_hours"))]
colnames(df)
```

## Top 10 variables by IV
```{r, echo=FALSE}
# Weight of Evidence -------------------------------------
woe_df <- df

# Response to numeric
woe_df$is_labor_union <- ifelse(woe_df$is_labor_union == 'Same', 1, 0)

# Create Info Tables
IV <- create_infotables(data=woe_df, y="is_labor_union")
print(IV$Summary$Variable[1:10])
```

## Test Model and Model Performance
```{r}
firstmodel <- glm(is_labor_union ~ industry_code + occupation_code + govt_employee + wage_grps + total_employed + age_grp + tax_status_grp + working_weeks_grp, data=train, family="binomial")
predictions <- predict(firstmodel, test, type="response")
predictions <- as.factor(predictions > 0.5)
actual <- test$is_labor_union
actual <- as.factor(actual == "Same")
confusionMatrix(table(actual,predictions))
```
## Synthetic Minority Oversampling (SMOTE)
```{r}
set.seed(1234)

balancedf <- smote(is_labor_union ~ ., data=df)
sample2 <- sample(c(TRUE, FALSE), nrow(df), replace=TRUE, prob=c(0.9,0.1))
balancetrain  <- balancedf[sample, ]
balancetest   <- balancedf[!sample, ]
```
```{r}
bestmodel <- glm(is_labor_union ~ industry_code + occupation_code + govt_employee + wage_grps + total_employed + age_grp + tax_status_grp + working_weeks_grp, data=balancetrain, family="binomial")
predictions <- predict(bestmodel, balancetest, type="response")
predictions <- as.factor(predictions > 0.5)
probabilities <- predict(bestmodel, balancetest, type="response")
actual <- balancetest$is_labor_union
actual <- as.factor(actual == "Same")
cm <- confusionMatrix(table(actual,predictions))
cm
```
## Plot Confusion Matrix
```{r}
ggplot(data.frame(cm$table),aes(actual,predictions)) +
  scale_fill_gradient(low = "white", high = "blue") +
  geom_tile(aes(fill= Freq)) +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(aes(label = sprintf("%d", Freq), color = Freq))
```

# ROC and AUC
```{r}
roc_obj <- roc(as.numeric(actual), as.numeric(probabilities))
plot(roc_obj, main="Receiver Operating Characteristic (ROC) Curve", colorize=FALSE)

# Calculate and print the AUC score
auc_score <- auc(roc_obj)
print(paste("AUC Score:", auc_score))
```